Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex
minor sections 1, 2

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}
\address[stpetersburg]{St. Petersburg State University, Russia.}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, with important consequences on the development effort
required and resulting system.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed
 by our experience in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years, as well as by lessons learned from several science
gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures
 that cover
the main patterns that we encountered in the various systems that we
studied during the last few years. The architectures are described
within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and 
developers.

@@

\label{sec:introduction-engines}
In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}.  Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow engines and
domain-specific pipeline systems.
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may require specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler, or
the presence of a file catalog. In the analysis presented hereafter,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

  including: authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires to deal with the

@@

  or a cloud, it may for instance be achieved using libraries that

@@

  such as: upload, download, deletion, browsing, replication, caching,

@@

  use it. The infrastructure might offer various data storage backends

@@

  engine, including: workflow submission, monitoring, termination,

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

>>>>

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Dept. Clinical Epidemiology, Biostatistics and Bioinformatics, Academic Medical Center, University of Amsterdam, NL.}
\address[stpetersburg]{St. Petersburg State University, Russia.}

@@

Workflow engines are critical for the efficient and transparent
exploitation of distributed infrastructures in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, with important consequences for the development effort
required and resulting system.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed
 by our experience in the development and
sustained operation of the CBRAIN~\cite{SHER-14},
VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc} science gateways for medical image analysis during
the past 7 years, as well as by lessons learned from several science
gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures
 that cover
the main patterns that we encountered in the various systems that we
studied during the last few years. The architectures are described
within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and 
developers.

@@

\label{sec:introduction-engines}
In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}.  Such workflow engines usually describe
applications using a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that coordinates and submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}), using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow engines and
domain-specific pipeline systems.
Some workflow engines, 
usually from the e-Science community, 
may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

An infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. The infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may require specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler, or
the presence of a file catalog. In the analysis presented hereafter,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

  including authentication, submission, monitoring, termination,
  deletion, etc. Controlling tasks requires dealing with the

@@

  or a cloud, control may be achieved for instance using libraries that

@@

  such as upload, download, deletion, browsing, replication, caching,

@@

  use them. The infrastructure might offer various data storage backends

@@

  engine, including workflow submission, monitoring, termination,

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where applications, users and other
resources are stored.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine,
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}
